Examples of Customization

### 2017-06-05 README.Rmd
Add Travis CI checking

---
output: github_document
---

@@

@@

```{r, echo = FALSE, message = FALSE}
library(eesim)
library(ggplot2)
library(dplyr)
library(tidyr)
library(viridisLite)
library(gridExtra)
```

@@

1. Generation of exposure data; 
2. Generation of outcome data;
3. Fitting models to simulated data; and 
4. Evaluating model performance on simulated data.

@@

The package creates time series that are relevant for environmental epidemiology studies of ambient exposures (e.g., studies of acute mortality risks associated with daily air pollution concentration, daily temperature, or occurance of a community-wide extreme event like a heat wave).

@@

The main function of this package is the `eesim` function. You can use the `eesim` function to conduct all four steps of the simulation process at once (simulate exposure data, simulate outcome data, fit models to simulated data, and evaluate model performance).

The `eesim` function requires [required inputs]. The function returns a list with two elements. The first element gives simulation-specific results: the estimated effect, standard error, t- and p-values, and upper and lower 95% confidence bounds when a model was applied to each set of simulated data. The second element gives some measures of model assessment, assessed over all simulations, including the mean beta and relative risk estimates across simulations.

For example, you can use the following call to (1) generate 10 observations of a continuous exposure with mean 100 and standard deviation 10, with a seasonal trend in expected value; (2) generate 10 associated outcome values, where the outcome has an average value of 20 and a relative risk of 1.10 per one-unit increase in the exposure; (3) fit a generalized linear model that controls for long-term and seasonal trends with a natural cubic spline with 1 degree of freedom per year; and (4) evaluate the performace of that model on the simulated data:

[To do-- increase `n` and `n_reps`, but wait until we've edited the vignette, because that will increase time to render. We could change `df` then, too (and maybe change to case-crossover? That may be easier for a first example)]

```{r, warning = FALSE, message = FALSE}
ex_sim <- eesim(n_reps = 100, n = 10, central = 100, sd = 10,
                exposure_type = "continuous", 
                average_outcome = 20, rr = 1.10,
                model = "spline", df_year = 1)
head(ex_sim[[1]])
ex_sim[[2]]
```

The first element of the returned object can be used to explore the behavior of individual simulations. For example, to look at the relative risk point estimate and 95% confidence interval from each of the 100 simulations, as well as which 95% confidence intervals include the true relative risk using the `coverage_plot` function that comes with the package:

```{r fig.width = 4, fig.height = 5, fig.align = "center"}
coverage_plot(ex_sim[[1]], true_param = 1.10)
```

The second element of the returned object gives the following overall summaries of model performance across all simulations:

```{r echo = FALSE}
eesim_overall_output <- data_frame(element = paste0("`",
                                                    colnames(ex_sim[[2]]),
                                                    "`"),
                                   desc = c("The mean estimated effect (log relative risk) over all simulations",
                                            "The mean relative risk over all simulations",
                                            "Variance of the point estimates of the estimated effect over all simulations",
                                            "The mean across simulations of the variances of each estimated effect",
                                            "Relative bias of the mean of the estimated coefficients",
                                            "Percent of simulations for which the 95% confidence interval estimate includes the true value",
                                            NA))
knitr::kable(eesim_overall_output, col.names = c("Variable", "Description"))
```

##Piece-by-piece breakdown of package utility

@@

The first task of the package is generating exposure data. This can be dones with the `sim_exposure` function. In this function, the user can specify whether they would like the exposure data to be binary or continuous (`exposure_type`). For continuous exposure data, the user can specify the mean (`central`) and standard deviation (`sd`) of the exposure data:

```{r}
x_cont <- sim_exposure(n = 1000, central = 50, sd = 5,
                      exposure_type = "continuous") 
x_cont %>% slice(1:5)
```
```{r fig.width = 4, fig.height = 2, fig.align = "center"}
ggplot(x_cont, aes(x = date, y = x)) + geom_point(alpha = 0.2)
```

You can plot a calendar plot of this simulated exposure time series using the `calendar_plot` function that comes with the package:

@@

For binary exposure data, the `central` argument gives the probability of exposure on a study day:

```{r}
x_bin <- sim_exposure(n = 1000, central = 0.1,
                      exposure_type = "binary")
x_bin %>% slice(1:5)
```

```{r fig.align = "center", fig.width = 8, fig.height = 2.5}
calendar_plot(x_bin, type = "discrete", labels = c("Not exposed", "Exposed"))
```

For environmental epidemiology applications, exposure data often has a seasonal trend and / or long-term trend. The default for `sim_exposure` is to simulate the exposure data without a time trend (`trend = "no trend"`). However, we have also built in several time trends from which a user can to choose to simulate exposure data with a time trend, either seasonal or long-term or both. These trend patterns differ slightly depending on whether the user is simulating binary or continuous data. [add reference for Schwartz paper where we got some of the trend equations] Below are plots of the built-in trends for continuous exposure data from which the user may choose.

```{r fig.width = 7, fig.height = 4, echo = FALSE, message = FALSE}
data.frame(day = 1:1000) %>%
  mutate(`"no trend"` = calc_t(n = 1000, trend = "no trend"),
         `"cos1"` = calc_t(n = 1000, trend = "cos1"),
         `"cos2"` = calc_t(n = 1000, trend = "cos2"),
         `"cos3"` = calc_t(n = 1000, trend = "cos3"),
         `"linear"` = calc_t(n = 1000, trend = "linear"),
         `"curvilinear"` = calc_t(n = 1000, trend = "curvilinear"),
         `"cos1linear"` = calc_t(n = 1000, trend = "cos1linear")) %>%
  gather(trend_method, trend_value, -day) %>%
  mutate(trend_method = factor(trend_method,
                               levels = c('"no trend"', 
                                          '"linear"',
                                          '"curvilinear"',
                                          '"cos1"',
                                          '"cos1linear"',
                                          '"cos2"',
                                          '"cos3"'))) %>%
  ggplot(aes(x = day, y = trend_value)) + 
  geom_line() + facet_wrap(~ trend_method, ncol = 4) + 
  theme_bw()
```

The trend options are similar for binary exposure, but exclude  "curvilinear" and "cos1linear" and include a "monthly" trend, for which the user enters a vector of 12 probabilities of exposure, one for each month.

Here is an example of generating continuous exposure data with a "cos1linear"" 
trend:

```{r, fig.width = 5, fig.height = 3, fig.align = "center"}
testexp <- sim_exposure(n=1000, central = 50, sd = 5, trend = "cos1linear",
                        amp = .6, exposure_type = "continuous")
head(testexp)
qplot(testexp$date, testexp$x)+geom_point()+coord_cartesian(ylim = c(0,110)) + 
  labs(title = "Exposure with a cos1linear trend", x = "Date", y="Exposure")+theme_bw()
```

```{r fig.align = "center", fig.width = 8, fig.height = 2.5}
calendar_plot(testexp, type = "continuous")
```

The default amplitude for the trend is .6, but we can adjust the "amp" parameter
for a smaller or larger trend amplitude:

```{r, fig.width = 5, fig.height = 3, fig.align = "center"}
smallamp <- sim_exposure(n=1000, central = 50, sd = 5, trend = "cos1linear",
                        amp = .2, exposure_type = "continuous")
qplot(smallamp$date, smallamp$x) +
  geom_point() + coord_cartesian(ylim = c(0,110)) +
  labs(title = "Cos1linear exposure with smaller amplitude", x="Date", y="Exposure") + theme_bw()
```

```{r fig.align = "center", fig.width = 8, fig.height = 2.5}
calendar_plot(smallamp, type = "continuous")
```

Here is an example of generating binary exposure data with a "monthly" trend 
starting from June 1, 2002.  The probabilities for the central parameter are for each month of the year, starting in January.

```{r, message = FALSE, fig.width = 5, fig.height = 3, fig.align = "center"}
testbin <- sim_exposure(n=1000, central = c(.1,.1,.2,.3,.4,.4,.5,.6,.5,.3,.2,.1),
                        trend = "monthly", exposure_type = "binary", 
                        start.date = "2002-06-01")
head(testbin)
qplot(testbin$date, testbin$x)+geom_point()+labs(title = "Binary exposure data with monthly trend", x="Date", y="Exposure")+theme_bw()
```

```{r fig.align = "center", fig.width = 8, fig.height = 2.5}
calendar_plot(testbin, type = "discrete", labels = c("Not exposed", "Exposed"))
```

The following figure shows the steps taken in simulating exposure ...

```{r echo = FALSE, fig.width = 8, fig.height = 3}
# binary_trend 
# binary_continuous
continuous_trend <- data_frame(t = calc_t(n = 1000, trend = "cos1")) %>%
  mutate(x = 100 * t,
         date = seq(from = as.Date("2000-01-01"),
                    by = 1, length.out = 1000)) %>%
  ggplot(aes(x = date, y = x)) + geom_point() + theme_minimal() + 
  ylim(c(0, 200))
continuous_simulated <- continuous_exposure(n = 1000, mu = 100,
                                            sd = 10, trend = "cos1") %>%
  ggplot(aes(x = date, y = x)) + geom_point() + theme_minimal() + 
  ylim(c(0, 200))
grid.arrange(continuous_trend, continuous_simulated, ncol = 2)
```

@@

Next, the eesim package uses the exposure data to generate health outcome data with or without seasonal trends. eesim creates "baseline" values for outcome based on the user-specified trend and average outcome. Next it uses the following function to relate exposure and relative risk to outcomes:
$$\lambda = e^{\log{(baseline)}+\log{(relative risk)}*exposure}$$
The values of lambda are then used to randomize the outcome values around the baseline with a Poisson($\lambda\$) distribution.

Here is an example of generating health outcome data with an upward linear trend using exposure data with a "cos1" trend:

```{r fig.width = 5, fig.height = 3, fig.align = "center"}
testexp2 <- sim_exposure(n=1000, central = 100, sd = 10, trend = "cos1", exposure_type = "continuous")
testout <- sim_outcome(exposure = testexp2, average_outcome = 22, trend = "linear", rr = 1.01)
qplot(testout$date, testout$outcome) + geom_point() +
  labs(title = "Health outcomes with a linear trend", x="Date", y="Outcome") + theme_bw()
```

```{r fig.align = "center", fig.width = 8, fig.height = 2.5}
calendar_plot(testout %>% select(date, outcome), type = "continuous")
```

###Fitting models

>>>>

---
output: github_document
bibliography: vignettes/eesim_bibfile.bib
---

@@

[![Travis-CI Build Status](https://travis-ci.org/sakoehler7/eesim.svg?branch=master)](https://travis-ci.org/sakoehler7/eesim)

@@

```{r, echo = FALSE, message = FALSE}
library(eesim)
library(ggplot2)
library(dplyr)
library(tidyr)
library(gridExtra)
```

@@

1. Generation of exposure data (simulated or from real data); 
2. Generation of simulated outcome data;
3. Fitting models to generated data; and 
4. Evaluating model performance on generated data.

@@

The package creates simulated time series data that are relevant for environmental epidemiology studies of ambient exposures (e.g., studies of acute mortality risks associated with daily air pollution concentration, daily temperature, or occurance of a community-wide extreme event like a heat wave). Simulated environmental datasets like those created by the package can be used in to assess the performance of statistical models meant to estimate the association between exposure level and outcome risk, to estimate power for a planned study, and to develop a better understanding of the data generating processes behind observed environmental datasets. Such time series are often characterized by both seasonal and long-term trends in both the exposure of interest and the outcome. For example, the following plot shows time series of daily ozone concentration (in parts per billion [ppb]) and cardiovascular deaths in Chicago, IL (1996--2000), with smoothed lines overlaid on the raw data to show patterns over time.

```{r message = FALSE, echo = FALSE, fig.width = 5, fig.align = "center"}
library(dlnm)
data("chicagoNMMAPS")
chicagoNMMAPS %>% 
  tbl_df() %>% 
  filter(year >= 1996) %>% 
  select(date, cvd, o3) %>% 
  rename(`Cardiovascular deaths` = cvd,
         `Ozone concentration (ppb)` = o3) %>% 
  gather(variable, value, -date) %>% 
  ggplot(aes(x = date, y = value)) + 
  geom_point(alpha = 0.5, size = 1) + 
  geom_smooth(se = FALSE, span = 0.1, method = "loess", color = "red") +
  facet_wrap(~ variable, ncol = 1, scales = "free_y") + 
  theme_classic() + 
  labs(x = "Date", y = "Variable value")
```

@@

The main function of this package is the `eesim` function. You can use the `eesim` function to conduct all four steps of the simulation process at once (generate exposure data, generate outcome data, fit models to simulated data, and evaluate model performance).

The `eesim` function requires inputs on:

- `n`: The desired number of observations per simulated dataset (for a daily time series, this is the desired number of days in the simulated dataset)
- `n_reps`: The desired number of simulated datasets
- `exposure_type`: Whether the exposure is binary (e.g., occurence of an extreme event like a heat wave or wildfire) or continuous (e.g., concentration of a pollutant)
- `rr`: The relative rate of the outcome associated with the exposure. For a binary exposure, this is the relative rate associated with the exposure compared to a similar day without the exposure. For a continuous exposure, this is the relative rate associated with a one-unit increase in the exposure. 
- `model`: The model to be used to estimate the association between exposure and outcome in the simulated datasets, either to estimate power of a planned analysis or to otherwise evaluate the performance (e.g., coverage, bias) of a model on the simulated datasets.

A number of optional inputs can also be specified, including arguments to adjust the shape of seasonal or long-term trends in the exposure or outcome data or custom arguments to use at different steps of the data generation.

The function returns a list with three elements. The first element is a list with all the simulated datasets. The second element gives simulation-specific results for each simulated dataset: the estimated effect, standard error, t- and p-values, and upper and lower 95% confidence bounds when a model was applied to each of the simulated datasets. The third element gives some measures of model assessment, assessed over all simulations, including the mean beta and relative risk estimates across simulations.

```{r message = FALSE, echo = FALSE, results='hide'}
chicagoNMMAPS %>% 
  tbl_df() %>% 
  filter(year >= 1996) %>% 
  select(date, cvd, o3) %>% 
  gather(variable, value, -date) %>% 
  group_by(variable) %>% 
  summarize(mean = mean(value, na.rm = TRUE),
            sd = sd(value, na.rm = TRUE))

library(splines)
a <- chicagoNMMAPS %>% 
  tbl_df() %>% 
  filter(year >= 1996) %>% select(date, o3)
mod <- lm(o3 ~ ns(scale(time, scale = FALSE, center = TRUE), 7 * 14), data = chicagoNMMAPS)
sd(residuals(mod))
```

For example, in the observed data from Chicago, IL, shown in the plots above, daily ozone concentrations have a mean of about 20 ppb and standard deviation of about 7 ppb after removing seasonal and long-term trends. The average number of cardiovascular deaths per day is around 50. Here is the code, and a plot of the resulting data, for generating a dataset with similar characteristics for use in a power analysis or to evaluate model performance (later in the vignette, we will show how to use customization to further improve the simulation of data for this example, including avoiding negative values of ozone concentration in simulated data):

```{r, warning = FALSE, message = FALSE}
sim_chicago <- create_sims(n_reps = 1, n = 365 * 5, central = 20, sd = 7,
                           exposure_type = "continuous", exposure_trend = "cos1",
                           exposure_amp = -.6, average_outcome = 50,
                           outcome_trend = "cos1", outcome_amp = 0.2, 
                           rr = 1.0005, start.date = "1996-01-01")
head(sim_chicago[[1]])
```

```{r echo = FALSE, message = FALSE, echo = FALSE, fig.width = 5, fig.align = "center"}
sim_chicago[[1]] %>% 
  rename(Exposure = x,
         Outcome = outcome) %>% 
  gather(variable, value, -date) %>% 
  mutate(variable = factor(variable, levels = c("Outcome", "Exposure"))) %>% 
  ggplot(aes(x = date, y = value)) + 
  geom_point(alpha = 0.5, size = 1) + 
  facet_wrap(~ variable, ncol = 1, scales = "free_y") + 
  theme_classic() + 
  geom_smooth(se = FALSE, method = "loess", color = "red", span = 0.1)
```

This simulated data can also be visualized using the `calendar_plot` function that comes with the package:

```{r fig.align = "center", fig.width = 8, fig.height = 7.5}
a <- calendar_plot(sim_chicago[[1]] %>% select(date, outcome), type = "continuous", 
                   legend_name = "Outcome") + 
  ggtitle("Outcome")
b <- calendar_plot(sim_chicago[[1]] %>% select(date, x), type = "continuous") + 
  ggtitle("Exposure")
grid.arrange(a, b, ncol = 1)
```

You can use the `eesim` function to generate multiple similar simulated datasets and investigate the performance of a specified model in estimating the association between ozone concentration and the risk of cardiovascular death in 20 simulated datasets. You must write a function with the code to fit the model you desire to fit to the simulated data (more details for writing this function are provided later in the vignette), and then you can use the `eesim` function to generate lots of simulated datasets, fit that model, and assess its performance using a call like:

```{r echo = FALSE}
set.seed(16)
```

```{r, warning = FALSE, message = FALSE}
ex_sim <- eesim(n_reps = 100, n = 365 * 5, central = 20, sd = 7,
                exposure_type = "continuous", exposure_trend = "cos1",
                exposure_amp = -.6, average_outcome = 50,
                outcome_trend = "cos1", outcome_amp = 0.2, 
                rr = 1.2, start.date = "1996-01-01",
                custom_model = spline_mod, custom_model_args = list(df_year = 7))
```

The `eesim` function returns a list with three elements:

```{r}
names(ex_sim)
```

The first element of the returned object is a list with all of the simulated datasets. For example, you can create a calendar plot of exposure in the first simulated dataset using the call:

```{r fig.height = 3, fig.width = 7, fig.align = "center"}
calendar_plot(ex_sim[["simulated_datasets"]][[1]] %>% select(date, x), type = "continuous")
```

The second element of the returned object gives the results of fitting the model to each of the simulated datasets. It can be used to explore the behavior of individual simulations:

```{r}
head(ex_sim[["indiv_performance"]])
```

After running the simulation, you can look at the relative risk point estimate and 95% confidence interval from each of the 100 simulations, as well as which 95% confidence intervals include the true relative rate, using the `coverage_plot` function that comes with the package:

```{r fig.width = 4, fig.height = 5, fig.align = "center"}
coverage_plot(ex_sim[["indiv_performance"]], true_param = 1.2)
```

The third element of the list returned by a call to `eesim` gives the following overall summaries of model performance across all simulations:

```{r echo = FALSE}
eesim_overall_output <- data_frame(element = paste0("`",
                                                    colnames(ex_sim[[3]]),
                                                    "`"),
                                   desc = c("**Mean estimate**: The mean of the estimated log relative rate over all simulations.",
                                            "**Mean estimated relative rate**: The mean of the estimated relative rate over all simulations.",
                                            "**Variance across estimates**: Variance of the point estimates (estimated log relative risk) over all simulations.",
                                            "**Mean variance of estimate**: The mean of the variances of the estimated effect (estimated log relative risk) across all simulations.",
                                            "**Relative bias**: Difference between the estimated log relative risk and true log relative risk as a proportion of the true log relative risk.",
                                            "**95% confidence inverval coverage**: Percent of simulations for which the 95% confidence interval estimate of log relative risk includes the true value of log relative risk.",
                                            "**Power**: Percent of simulations for which the null hypothesis that the log relative risk equals zero is rejected based on a p-value of 0.05."))
knitr::kable(eesim_overall_output, col.names = c("Variable", "Description"))
```

For example, here are the overall results for the simulation fit above:

```{r}
ex_sim[["overall_performance"]]
```

In later sections of this vignette, we will show how to customize steps in the generation of the simulated data to further improve this example simulation.

As another basic example, here is a plot of the dates of extreme heat days (defined as a day with temperature at or above the 98\textsuperscript{th} percentile temperature in Chicago between 1987 and 2000) in the observed Chicago dataset (points are jittered along the y-axis to limit overlapping):

```{r echo = FALSE, message = FALSE, echo = FALSE, fig.width = 5, fig.align = "center", fig.height = 2}
chicagoNMMAPS %>% 
  mutate(temp = temp >= quantile(temp, probs = 0.98)) %>% 
  tbl_df() %>% 
  filter(year >= 1996) %>% 
  select(date, temp) %>% 
  rename(`Extreme heat day` = temp) %>% 
  gather(variable, value, -date) %>% 
  ggplot(aes(x = date, y = value)) + 
  geom_jitter(alpha = 0.5, size = 0.7, fill = NA, width = 0, height = 0.1) + 
  facet_wrap(~ variable, ncol = 1, scales = "free_y") + 
  theme_classic() + 
  labs(x = "Date", y = "Variable value")
```

```{r echo = FALSE, results="hide"}
chicagoNMMAPS %>% 
  mutate(temp = temp >= quantile(temp, probs = 0.98)) %>% 
  tbl_df() %>% 
  filter(year >= 1996) %>% 
  group_by(month) %>% 
  summarize(p_heat = mean(temp))
```

In this observed data, there is (unsurprisingly) a strong seasonal trend in this binary exposure of extreme heat days. The percent of days that are extreme heat days is 0% for all months expect June (about 5% of days in observed data were extreme heat days), July (about 12% of days), and August (about 2% of days). Similar exposure time series can be simulated with the call:

```{r}
sim_chicago2 <- create_sims(n_reps = 1, n = 365 * 5, sd = 1,
                            central = c(0, 0, 0, 0, 0, 0.05, 0.12, 0.02, 0, 0, 0, 0),
                            exposure_type = "binary", exposure_trend = "monthly",
                            exposure_amp = -.6, average_outcome = 50,
                            outcome_trend = "cos1", outcome_amp = 0.2, 
                            rr = 1.05, start.date = "1996-01-01")
```

Here is an example of the simulated exposure data:

```{r echo = FALSE, message = FALSE, echo = FALSE, fig.width = 5, fig.align = "center", fig.height = 2}
sim_chicago2[[1]] %>% 
  select(date, x) %>% 
  rename(`Extreme heat day` = x) %>% 
  gather(variable, value, -date) %>% 
  ggplot(aes(x = date, y = value)) + 
  geom_jitter(alpha = 0.5, size = 0.7, fill = NA, width = 0, height = 0.1) + 
  facet_wrap(~ variable, ncol = 1, scales = "free_y") + 
  theme_classic() + 
  labs(x = "Date", y = "Variable value")
```

Again, both the observed and simulated exposure data can also be plotted using the `calendar_plot` function:

```{r fig.align = "center", fig.width = 8, fig.height = 7}
a <- chicagoNMMAPS %>% 
  mutate(temp = temp >= quantile(temp, probs = 0.98)) %>% 
  tbl_df() %>% 
  filter(year >= 1996) %>% 
  select(date, temp) %>% 
  calendar_plot(type = "discrete", labels = c("Extreme heat day", "Other day")) + 
  ggtitle("Observed exposure data")
b <- sim_chicago2[[1]] %>% 
  select(date, x) %>% 
  calendar_plot(type = "discrete", labels = c("Extreme heat day", "Other day")) + 
  ggtitle("Simulated exposure data")
grid.arrange(a, b, ncol = 1)
```

The comparison of the observed and simulated data in this case suggests some clustering in the observed data that is not evident in the simulated data, suggesting that the probability of exposure may be higher on a day near other extreme heat days.

The `eesim` function can be used to assess the performance of a GLM in estimating relative risk of cardiovascular mortality for extreme heat days compared to other days using:

```{r echo = FALSE}
set.seed(2416)
```

```{r}
ex_sim2 <- eesim(n_reps = 100, n = 365 * 5, 
                 central = c(0, 0, 0, 0, 0, 0.05, 0.12, 0.02, 0, 0, 0, 0),
                 exposure_type = "binary", exposure_trend = "monthly",
                 exposure_amp = -.6, average_outcome = 50,
                 outcome_trend = "cos1", outcome_amp = 0.2, 
                 rr = 1.05, start.date = "1996-01-01",
                 custom_model = spline_mod, custom_model_args = list(df_year = 7))
```

As before, a plot of CI coverage can be created with `coverage_plot`:

```{r fig.width = 4, fig.height = 5, fig.align = "center"}
coverage_plot(ex_sim2[["indiv_performance"]], true_param = 1.05)
```

Here are the overall estimates in this case for model performance:

```{r}
ex_sim2[["overall_performance"]]
```

In this case, the expected power is low.

The `power_calc` function in the package allows you to extend on this simulation functionality to create power curves for an analysis given an anticipated underlying process of data generation. This function will create simulations for several different values of number of days in the study (`n`), average daily outcome counts (`average_outcome`), or expected association between exposure and outcome (`rr`).

For example, the following call generates a power curve that explores how expected power changes with an increasing number of days for the heat wave analysis example just presented (as a warning, this call takes a few minutes to run, since it's simulating many datasets):

```{r fig.width = 5, fig.height = 3.5, fig.align = "center"}
ex_power_calc <- power_calc(varying = "n", values = floor(365.25 * seq(1, 20, by = 5)),
                            n_reps = 100, rr = 1.05,
                            central = c(0, 0, 0, 0, 0, 0.05, 0.12, 0.02, 0, 0, 0, 0),
                            exposure_type = "binary", exposure_trend = "monthly", 
                            exposure_amp = -.6, average_outcome = 50,
                            outcome_trend = "cos1", outcome_amp = 0.2, 
                            custom_model = spline_mod, custom_model_args = list(df_year = 7),
                            plot = FALSE)
ex_power_calc %>% 
  ggplot(aes(x = values, y = power)) + 
  geom_line() + 
  ylim(0, 1) + 
  labs(x = "Number of days in the study", y = "Power") + 
  theme_bw()
```

## Piece-by-piece breakdown of package utility

@@

The first task of the package is generating exposure data. This can be done with the `sim_exposure` function. In this function, the user can specify whether he or she would like to generate exposure data that is binary or continuous (`exposure_type`). For continuous exposure data, the user must specify the mean (`central`) and standard deviation (`sd`) of the exposure data. For example, the following call simulates a dataframe of exposure data for an exposure that is normally distributed, with a mean value of 50, a standard deviation of 5, and no long-term or seasonal trends:

```{r}
x_cont <- sim_exposure(n = 1000, central = 50, sd = 5, exposure_type = "continuous") 
x_cont %>% slice(1:5)
```
```{r fig.width = 4, fig.height = 2, fig.align = "center"}
ggplot(x_cont, aes(x = date, y = x)) + geom_point(alpha = 0.2) + 
  theme_classic()
```

You can plot a calendar plot of this simulated exposure time series using the `calendar_plot` function that comes with the package. Within this function, the type of data ("continuous" or "discrete") must be specified:

@@

You can similarly use the `sim_exposure` function to simulate a binary exposure (e.g., occurence of an extreme event). For binary exposure data, the `central` argument of `sim_exposure` must also be expressed, but in this case it gives the probability of exposure on a study day:

```{r}
x_bin <- sim_exposure(n = 1000, central = 0.05, exposure_type = "binary")
x_bin %>% slice(1:5)
```

Again, the `calendar_plot` function can be used to visualize the generated time series. In the case of binary exposure data, the labels to be used in the legend for each outcome level must also be specified using the `labels` argument:

```{r fig.align = "center", fig.width = 8, fig.height = 2.25}
calendar_plot(x_bin, type = "discrete", labels = c("Not exposed", "Exposed"))
```

So far, these `sim_exposure` calls have been used to simulate basic exposure data, without long-term or seasonal trends. However, for environmental epidemiology applications, exposure data often has a seasonal trend and / or long-term trend, and these temporal trends can serve as confounders in assessing the association between time-varying environmental exposures and health outcomes. The `sim_exposure` function therefore includes options to generate exposure data with long-term and seasonal trends relevant to environmental time series studies, through the `trend` argument.

The default for `sim_exposure` is to simulate the exposure data without a time trend (`trend = "no trend"`). However, we have also built in several time trends from which a user can to choose to simulate exposure data with a time trend, either seasonal or long-term or both, based on trend patterns used in a simulation study of case-crossover studies as a method of controlling for seasonal and long-term trends in environmental epidemiology studies [@Bateson1999]. These trend patterns differ slightly depending on whether the user is simulating binary or continuous data. Below are plots of the built-in trends for continuous exposure data from which the user may choose. 
```{r fig.width = 7, fig.height = 4, echo = FALSE, message = FALSE}
data.frame(day = 1:1000) %>%
  mutate(`"no trend"` = calc_t(n = 1000, trend = "no trend"),
         `"cos1"` = calc_t(n = 1000, trend = "cos1"),
         `"cos2"` = calc_t(n = 1000, trend = "cos2"),
         `"cos3"` = calc_t(n = 1000, trend = "cos3"),
         `"linear"` = calc_t(n = 1000, trend = "linear"),
         `"curvilinear"` = calc_t(n = 1000, trend = "curvilinear"),
         `"cos1linear"` = calc_t(n = 1000, trend = "cos1linear")) %>%
  gather(trend_method, trend_value, -day) %>%
  mutate(trend_method = factor(trend_method,
                               levels = c('"no trend"', 
                                          '"linear"',
                                          '"curvilinear"',
                                          '"cos1"',
                                          '"cos1linear"',
                                          '"cos2"',
                                          '"cos3"'))) %>%
  ggplot(aes(x = day, y = trend_value)) + 
  geom_line() + facet_wrap(~ trend_method, ncol = 4) + 
  theme_bw()
```

You can use the `amp` argument to adjust the seasonal trend in any of the patterns with a seasonal trend. For example, here are plots of trends using `tren = "cos1linear"` for different  values of `amp`:

```{r fig.width = 7, fig.height = 2, echo = FALSE, message = FALSE}
data.frame(day = 1:1000) %>%
  mutate(`"amp = 0.5"` = calc_t(n = 1000, trend = "cos1linear", amp = 0.5),
         `"amp = 0.1"` = calc_t(n = 1000, trend = "cos1linear", amp = 0.1),
         `"amp = 0.9"` = calc_t(n = 1000, trend = "cos1linear", amp = 0.9),
         `"amp = -0.5"` = calc_t(n = 1000, trend = "cos1linear", amp = -0.5)) %>%
  gather(trend_method, trend_value, -day) %>%
  ggplot(aes(x = day, y = trend_value)) + 
  geom_line() + facet_wrap(~ trend_method, ncol = 4) + 
  theme_bw()
```

The long-term trend in expected values can be changed in a similar way with the `exposure_trend` argument in `eesim`.

Here is an example of generating continuous exposure data with a "cos1linear" trend for an exposure with a mean value of 50 and a standard deviation of 10:

```{r, fig.width = 5.5, fig.height = 5, fig.align = "center"}
testexp <- sim_exposure(n = 365 * 3, central = 50, sd = 10, trend = "cos1linear",
                        exposure_type = "continuous")
a <- ggplot(testexp, aes(x = date, y = x)) +  
  geom_point(alpha = 0.5, size = 0.8) + 
  coord_cartesian(ylim = c(0,110)) + 
  labs(title = "Exposure with a 'cos1linear' trend", x = "Date", y="Exposure") + 
  theme_classic()
b <- calendar_plot(testexp, type = "continuous") + 
  ggtitle("Calendar plot of simulated exposure data") + 
  theme(legend.position = "bottom")
grid.arrange(a, b, ncol = 1)
```

Here is an example of changing the seasonal trend by changing the value for `amp` (the default value is 0.6) to simulate exposure data for an exposure with a smaller seasonal trend and with higher exposures typical in the summer than the winter:

```{r, fig.width = 5.5, fig.height = 5,  fig.align = "center"}
small_amp <- sim_exposure(n = 365 * 3, central = 50, sd = 10, trend = "cos1linear",
                        amp = -0.3, exposure_type = "continuous")
a <- ggplot(small_amp, aes(x = date, y = x)) +  
  geom_point(alpha = 0.5, size = 0.8) + 
  coord_cartesian(ylim = c(0,110)) + 
  labs(title = "Exposure with a 'cos1linear' trend", x = "Date", y="Exposure") + 
  theme_classic()
b <- calendar_plot(small_amp, type = "continuous") + 
  ggtitle("Calendar plot of simulated exposure data") + 
  theme(legend.position = "bottom")
grid.arrange(a, b, ncol = 1)
```

The trend options are similar for binary exposure, but exclude  "curvilinear" and "cos1linear". Further, binary exposures can also be simulated using a "monthly" trend (`trend = "monthly"`), in which the probability of exposure can vary by month. When using this "monthly" trend option, the `cental` argument to `sim_exposure` should include a vector with 12 separate probabilities (the first is for January, the second for February, etc.) rather than a single probability. Here is an example of generating binary exposure data with a monthly trend, starting from June 1, 2002, with higher probability of the exposure in summer months than in winter months:

```{r, message = FALSE, fig.width = 5.5, fig.height = 5,  fig.align = "center"}
testbin <- sim_exposure(n=1000, central = c(.05, .05, .1, .2, .4, .4, .5, .7, .5, .2, .1, .05),
                        trend = "monthly", exposure_type = "binary", 
                        start.date = "2002-06-01")
a <- testbin %>% 
  mutate(x = factor(x, levels = c(0, 1), labels = c("Not exposed", "Exposed"))) %>% 
  ggplot(aes(x = date, y = x)) + 
  geom_jitter(alpha = 0.5, size = 0.7, fill = NA, width = 0, height = 0.1) + 
  theme_classic() + 
  labs(x = "Date", y = "Exposure")
b <- calendar_plot(testbin, type = "discrete", labels = c("Not exposed", "Exposed")) + 
  ggtitle("Calendar plot of simulated exposure data") + 
  theme(legend.position = "bottom")
grid.arrange(a, b, ncol = 1)
```

The `sim_exposure` function works by first calculating the expected exposure on any date in the simulated time series (figure below, left). This expected value is a mean for a continuous exposure and a probability for a binary exposure. The `sim_exposure` function then draws random values from the appropriate distribution (normal distribution for a continuous exposure, binomial distribution for a binary exposure) based on this day-specific expected exposure value and, in the case of continuous exposure, the standard deviation of the exposure (figure below, right). For continuous exposure data, the standard deviation specified in the call to `eesim` should measure the standard deviation of each point from its expected value (i.e., from the expected line shown on the left below), not the overall standard deviation of exposure values across all days in the simulated data.

```{r echo = FALSE, fig.width = 8, fig.height = 3}
continuous_trend <- data_frame(t = calc_t(n = 1000, trend = "cos1")) %>% 
  mutate(x = 100 * t,
         date = seq(from = as.Date("2000-01-01"),
                    by = 1, length.out = 1000)) 
a <- continuous_trend %>%
  ggplot(aes(x = date, y = x)) + geom_line(color = "red") + theme_classic() + 
  ylim(c(0, 200))

continuous_simulated <- continuous_exposure(n = 1000, mu = 100,
                                            sd = 10, trend = "cos1") 
b <- continuous_simulated %>%
  ggplot(aes(x = date, y = x)) + geom_point() + 
  geom_line(data = continuous_trend, color = "red") + theme_classic() + 
  ylim(c(0, 200))

grid.arrange(a, b, ncol = 2)
```

Later in this vignette, we show how you can further customize this step of generating exposure data through the use of a user-created function, allowing extensive further flexibility in simulating exposure data.

@@

Next, the `sim_outcome` function simulates outcome data. The health data can have an underlying seasonal and / or long term trend in its baseline value, and then that baseline is adjusted for the risk associated with exposure, based on the generated exposure data for that day. The baseline outcome count for a given day ($B_t$) are based on a user-specified trend and user-specified average outcome per day over the simulated time period. Further, the expected outcome count on a given day is adjusted for exposure-related risk through a user-specified relative rate per unit increase in exposure ($RR$) and the simulated exposure for that day ($X_t$). The `eesim` function then uses the following equation to calculate the expected outcome count ($\lambda$) on a given day in the simulated time series, based on the expected baseline rate and exposure-related risk for that day:

$$
\log(\lambda_t) = \log{(B_t)}+\log{(RR)}*X_t
$$

For a binary outcome, the baseline count on a given day ($B_t$) is the expected outcome count for the day if there is not an event (e.g., in a heat wave study, a non-heat wave day). For a continuous exposure, the baseline count on a given day ($B_t$) is the expected outcome count for the day if exposure is at its mean value.

Once the expected count ($\lambda_t$) on each day of the simulated time series is calculated using this equation, the simulated count on each day is drawn as a random variable from a Poisson distribution with mean $\lambda_t$. Later we describe how customization can be used to simulate output counts in other ways.

Here is an example of generating health outcome data with an upward linear trend using exposure data with a "cos1" trend. In this case, there is a steady increase in the baseline outcome count over time, as well as a seasonal trend linked to the risk associated with the seasonally-varying exposure:
